---
title: "Practical ML Final Project"
author: "MTDUGDA"
date: "December 26, 2016"
output: html_document
---

##The Problem and the Goal of the Project

One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, our goal is to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. The goal of the project is to predict the manner in which the participants did the exercise. This is the "classe" variable in the training set.

#Instructions:
You may use any of the other variables to predict with. We should create a report describing how we built our model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.

##Background

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

##Data

The training data for this project are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:

https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv


The data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r,echo=TRUE}
library(caret)
library(Hmisc)
library(randomForest)
library(foreach)
library(doParallel)
library(rpart)
library(rpart.plot)
library(RColorBrewer)
library(rattle)
set.seed(5000)
```

##Getting the data
The training and testing data sets are obtained from the following URLs:

```{r,echo=TRUE}
trainUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "http://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

##Let's load data to memory
training <- read.csv(url(trainUrl), na.strings=c("NA","#DIV/0!",""))
testing <- read.csv(url(testUrl), na.strings=c("NA","#DIV/0!",""))
```

# partion the Data set
Let's partion the Training data set into two data sets: 60% for Training and 40% for Testing:
```{r,echo=TRUE}
inTrain <- createDataPartition(y=training$classe, p=0.6, list=FALSE)
myTraining <- training[inTrain, ]; myTesting <- training[-inTrain, ]
dim(myTraining); dim(myTesting)
```

##Cleaning the data
The following steps were used to clean the data:
##Step 1: 
Cleaning NearZeroVariance (NZV) Variables Run this code to view possible NZV Variables:

```{r,echo=TRUE}
myDataNZV <- nearZeroVar(myTraining, saveMetrics=TRUE)
#Run this code to create another subset without NZV #variables:
myNZVvars <- names(myTraining) %in% c("new_window", "kurtosis_roll_belt", "kurtosis_picth_belt",
"kurtosis_yaw_belt", "skewness_roll_belt", "skewness_roll_belt.1", "skewness_yaw_belt",
"max_yaw_belt", "min_yaw_belt", "amplitude_yaw_belt", "avg_roll_arm", "stddev_roll_arm",
"var_roll_arm", "avg_pitch_arm", "stddev_pitch_arm", "var_pitch_arm", "avg_yaw_arm",
"stddev_yaw_arm", "var_yaw_arm", "kurtosis_roll_arm", "kurtosis_picth_arm",
"kurtosis_yaw_arm", "skewness_roll_arm", "skewness_pitch_arm", "skewness_yaw_arm",
"max_roll_arm", "min_roll_arm", "min_pitch_arm", "amplitude_roll_arm", "amplitude_pitch_arm",
"kurtosis_roll_dumbbell", "kurtosis_picth_dumbbell", "kurtosis_yaw_dumbbell", "skewness_roll_dumbbell",
"skewness_pitch_dumbbell", "skewness_yaw_dumbbell", "max_yaw_dumbbell", "min_yaw_dumbbell",
"amplitude_yaw_dumbbell", "kurtosis_roll_forearm", "kurtosis_picth_forearm", "kurtosis_yaw_forearm",
"skewness_roll_forearm", "skewness_pitch_forearm", "skewness_yaw_forearm", "max_roll_forearm",
"max_yaw_forearm", "min_roll_forearm", "min_yaw_forearm", "amplitude_roll_forearm",
"amplitude_yaw_forearm", "avg_roll_forearm", "stddev_roll_forearm", "var_roll_forearm",
"avg_pitch_forearm", "stddev_pitch_forearm", "var_pitch_forearm", "avg_yaw_forearm",
"stddev_yaw_forearm", "var_yaw_forearm")
myTraining <- myTraining[!myNZVvars]
#To check the new size of observations
dim(myTraining)
```

# Cleaning Step 2: 
Removing first column of Dataset ID, i.e. Removing first ID variable:

```{r,echo=TRUE}
myTraining <- myTraining[c(-1)]
```

# Cleaning Step 3: 
Cleaning Variables with too many NAs:

```{r,echo=TRUE}
trainingV3 <- myTraining #creating another subset to iterate in loop
for(i in 1:length(myTraining)) { #for every column in the training dataset
if( sum( is.na( myTraining[, i] ) ) /nrow(myTraining) >= .6 ) { #if n?? NAs > 60% of total observations
for(j in 1:length(trainingV3)) {
if( length( grep(names(myTraining[i]), names(trainingV3)[j]) ) ==1) { #if the columns are the same:
trainingV3 <- trainingV3[ , -j] #Remove that column
}
}
}
}
#check the size of observations
dim(trainingV3)

#Seting back to our set:
myTraining <- trainingV3
rm(trainingV3)
#Now let us do the exact same 3 transformations but our #myTesting and testing data sets.
clean1 <- colnames(myTraining)
clean2 <- colnames(myTraining[, -58]) #already with classe column removed
myTesting <- myTesting[clean1]
testing <- testing[clean2]
#To check the  N# ofmyTesting observations
dim(myTesting)
#To check the new N# of testing observations
dim(testing)
```


```{r,echo=TRUE}
for (i in 1:length(testing) ) {
   for(j in 1:length(myTraining)) {
      if( length( grep(names(myTraining[i]),           names(testing)[j]) ) ==1) {
class(testing[j]) <- class(myTraining[i])
}
}
}

#And to make sure Coertion really worked:
testing <- rbind(myTraining[2, -58] , testing) 
#note row 2 does not mean anything, this will be removed here:
testing <- testing[-1,]

#Using Decision Tree ML algorithms for prediction: 
#treeModel <- rpart(classe ~ ., data=subTraining, method="class")
#predictionTree <- predict(treeModel, subTesting, type = "class")
#rpart.plot(treeModel, main="Classification Tree", extra=102, under=TRUE, faclen=0)
##
modFitA <- rpart(classe ~ ., data=myTraining, method="class")
#Note: to view the decision tree with fancy run this command:
fancyRpartPlot(modFitA)
#Predicting:
predictionsA <- predict(modFitA, myTesting, type = "class")
confusionMatrix(predictionsA, myTesting$classe)
```

#Using RandomForests ML algorithms for prediction: 
```{r,echo=TRUE}
modFitB <- randomForest(classe ~. , data=myTraining)
predictionsB <- predict(modFitB, myTesting, type = "class")
#(Moment of truth) Using confusion Matrix to test results:
confusionMatrix(predictionsB, myTesting$classe)
```

#Generating Files for Answering the Raised Questions:

Since RandomForests provides better results, we use the following function for better prediction:

```{r,echo=TRUE}
predictionsB <- predict(modFitB, testing, type = "class")
#Function to generate files with predictions to submit for assignment
pml_write_files = function(x){
n = length(x)
for(i in 1:n){
filename = paste0("problem_id_",i,".txt")
write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
}
}
pml_write_files(predictionsB)
```
